---
title: 'UTSSRP Tutorial: Bayesian Linear Regression'
author: "Prof. Gwen Eadie"
date: "2025-06-11"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
set.seed(1234)
# install.packages("visreg")
```

## Fitting a simple regression line to mock data (frequentist approach)

This exercise follows section 6.2 in Gelman, Hill, and Vehtari (2021).

1.  Simulate 20 fake data points from the model $$y_i = a +bx_i + \epsilon_i$$ where there is one predictor (covariate) that takes on values between 1 and 20, and where the intercept is 0.2 and the slope is 0.3, and $$\epsilon_i \sim N(0, \sigma=0.5)$$

```{r}
# write code and a function to simulate y values here. 
n = 20
x = 1:20

# intercept
beta0 = 0.2

# slope
beta1 = 0.3 

# sigma for random error
mysigma = rnorm(20, 0, 0.5) 
# simulated y values
y = beta0 + beta1*x + mysigma
```

2.  Fit a standard linear regression model to these fake data. You will need to create a `data.frame` containing both the predictor and the outcome. Use the function `lm` to fit the simple linear regression.

```{r}
# create data frame with covariate and outcome in columns
fakedata = data.frame(x, y)

# you should use the lm function and assign the results to a new object
lmfit = lm(y ~ x, data = fakedata)
# lmfit
```

3.  Now display the results using `print`. What does print show you? 

```{r}
print(lmfit)
```


4. Display results using `summary`, which gives a little more information (e.g., the standard error of each estimate)
```{r}
summary(lmfit)

```


5. Plot the data, regression line, and confidence bands
```{r}
# simple plot without confidence bands

# can use the visreg package to easily make fit line with confidence bands
# install.packages("visreg")
library(visreg)
visreg(lmfit)

```


## Inferring the parameters for a line (Bayesian approach)


Using the same mock data, infer the parameters of the intercept and slope. 
```{r}
# load packages
library(ggplot2)
library(tidyverse)
#install.packages("tidybayes")
library(tidybayes)
#install.packages("brms")
library(brms)
```

1.  Run the bayesian analysis using the `brm` function from the `brms` package. We will use the default priors in stan_lm. To read the help, do `?brm`.
```{r}
# use Bayesian Regression Model (brm) function from brms package
# Fit Bayesian linear model with default priors
bayes_fit <- brm(
  formula = y ~ x,
  data = fakedata)

```

```{r}
get_prior(y ~ x, data = fakedata)

```

2.  Look at the chains and summary of the output of the inferred posterior distribution
```{r}
# plot the Markov chains and marginal distribution of the draws
plot(bayes_fit)

# get the summary output
summary(bayes_fit)
```

3.  Plot the inferred relationship using the posterior distribution
```{r}
# Create a new data frame that will be used to generate predictions given the (Bayesian) regression prediction of y at a bunch x's
new_data <- data.frame(x = seq(1, 20, length.out = 100))

# Get fitted values for the new_data object
fitted_values <- fitted(bayes_fit, newdata = new_data)

# Use the posterior draws and new_data object to calculate credible intervals. Use the `posterior_epred` function for this
posterior_draws <- posterior_epred(bayes_fit, newdata = new_data)
    
# Calculate the credible interval (e.g., 95% CI) from the posterior draws using the quantile function:
credible_interval <- apply(posterior_draws, 2, function(draws_at_x) {
  c(mean = mean(draws_at_x),
    lower = quantile(draws_at_x, 0.025),
    upper = quantile(draws_at_x, 0.975))
})
# Now plot the line and credible intervals (I recommend using the ggplot2 package)

# Convert to data frame and add x values
credible_interval <- as.data.frame(t(credible_interval))
colnames(credible_interval) <- c("mean", "lower", "upper")
credible_interval$x <- new_data$x

library(ggplot2)

ggplot(credible_interval, aes(x = x, y = mean)) +
  geom_line(color = "blue", size = 1.2) +  # posterior mean line
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "blue", alpha = 0.2) +  # 95% credible band
  geom_point(data = fakedata, aes(x = x, y = y), color = "black", size = 1.5) +  # original data points
  labs(
    title = "Bayesian Regression with 95% Credible Interval",
    x = "x",
    y = "Predicted y"
  ) +
  theme_minimal()

```

