---
title: "m&m's exercise: Posterior Predictive Checking"
author: "Prof. Gwen Eadie"
date: "11/06/2025"
output: html_document
---


### Reminder of set-up for m\&m's exercise
$$P(\theta|y) = \frac{P(y|\theta)P(\theta)}{P(y)}$$

We defined:

-  $\theta$ as the percentage of blue candies made at the factory
-  $y$ as the number of blue candies observed in our data

The likelihood for our model was the binomial distribution with parameter $\theta$:

$$p(y|\theta) \propto \theta^{y}(1-\theta)^{n-y}.$$

This is our likelihood for $y$ blue candies, given the percentage $\theta$ produced at the factory.

We used the conjugate prior to the binomial distribution --- the beta distribution --- to quantify your prior information on the percentage $\theta$ of blue candies made at the factory:
$$p(\theta) \propto \theta^{\alpha-1}(1-\theta)^{\beta-1}$$
We chose some hyperparameter values
```{r}
# my hyperprior parameters for alpha and beta
# mya = 
# myb = 
```

Next, we collected our data, both individually, and as a class.
```{r}
# record your data before eating it!

# total number of trials per bag
# n = 15

# number of blue m&m's
# y =  # number of blue

# class data, sum of number of successes from each bag
# mydata = 

# number of bags

```


We calculated the posterior given your data and your prior distribution, and plotted it. Here is the code I wrote (commented out) if you want to use it.
```{r}
# set margins
par( mar = c(5,5,2,2))

# plot the posterior distribution
# curve( dbeta( x, shape1 = (mydata + mya), shape2 = (n-mydata+myb)),
#       xlab = expression(theta), ylab = expression(p(theta)) , 
#       main = "Posterior distribution for the probability of drawing a blue candy", 
#       cex.lab=2, lwd=2)
# 
# grid()


# add the prior distribution to the plot, to compare
# curve(dbeta(x, shape1 = mya, shape2 = myb),lwd=2, lty=3, add=TRUE, col="darkgrey")


# add a legend
# legend("topright", legend = c("posterior",                              as.expression(bquote("prior, "~alpha == .(mya)~"and"~beta == .(myb))) ),   lty=c(1,3), lwd=2, col=c("black", "darkgrey"))
```

### Exercise

Estimate the posterior predictive distribution and compare it to the distribution of successes (blue m&m's) from the class 

To calculate the posterior predictive distribution, follow these steps:

 *  draw a random value of $\theta$ from the posterior
 *  draw a $y$ (number of blue m&m's) from the likelihood (binomial distribution) given that value of $\theta$
 *  Repeat this many times.
 

```{r}
## simulte data from 500 students

# number of students/simulated observations
# nsims=

# object to store simulated observations
# yfuture = 

# draw a random theta given the posterior distribution we found in class
# thetatemp <- rbeta(......) 
  
# for each theta, draw a random y (number of blues) and store that. This should result in 500 observations of y (number of blues)
# yfuture <- rbinom(.....) 
```


Next, qualitatively compare the predicted distribution of successes (number of blues) to the distribution of successes for the whole class.

```{r}
# plot to compare real data to mock data

```






